\plaintitle

### 2016-04-23 paper/paper.tex
update niet compleet :-(

\section{Related Work}\label{sec:literature}

\section{Visualization}\label{sec:visualization}
The visualization consists of three parts:
\begin{itemize}
    \item The bubble view: a compact view on the play-offs per season
    \item The statistics (zoom) view: a more detailed view on a statistic of a selected team
    \item The team view: a detailed view on how good or bad a team scores on a specific field position
\end{itemize}
All three are discussed in more detail below. The three views are available to
the user on a single webpage. The user start on the bubble view and is able to
navigate to the statistics and team views by scrolling down or clicking on
teams.
%The webpage is built in such a way that it gives a hint to the user at the bottom
%or the top of a page when another view is available. By showing a different 
%background color and an arrow, we invite the visitor to explore other views. The 
%user is also at liberty to scroll through these views. However, a fixed scroll is
%implemeted so users will only end in complete views when moving around. Note that 
%the user also will be guided through all 3 views when performing actions described 
%in the section below.

\subsection{The bubble view}
This view supposes to inform the user of the results of the NBA play-offs per 
season. A circle is used to represent a team, the size its SRS score using a '?'
distribution, a stroke to illustrate it's region or medal won during regional 
competition and curved lines to inform the user which teams played against each 
other. The user has a timeline at its disposal to scroll through time. Note that
the team can be recognized by the logo inside the circle. Our initial desing filled
the circle with the red or blue depending on the region where the teams compete. 
However, the shade of colour was darker if their final ranking was higher in their
region. We've replaced this with the logo's of the teams as we thought this is more
user friendly and the extra information of the ranking doesn't outweight.
<picture of the bubble view compared with double elimination bracket>
A more popular visualization of a play-off is the team double elimination bracket.
Although this is a more common representation, we wanted to create a more compact
version by eliminating the recurring representation of a team. We also considered 
enclosed circles and a sunburst.
<figure of 2 alternatives https://thetuftersblog.wordpress.com/2016/03/07/team-ranking-play-off-infovis/>
Although the enclosed circles is a compact representation as well, we have continued 
with our own design as we believe this view is more original and innovative. We 
didn't encouter a similar view when searching for alternatives to visualize a 
play-off or championship. 
Our initial bubble view was too compact. It connected teams that played against
each other directly. When searching for alternatives to reflect on our choice we
discovered a similar visualization in a totally different context. 
<https://source.opennews.org/en-US/articles/nyts-512-paths-white-house/>
<comparison of original bubble view with white house paths>
This visualization gives a cleaner view of the competitors for a specific team
compared to our original sketch. Hence we decided to adapt our visualization. We
added an intermediate step between two teams to clearer indicate how teams competed
to become NBA champion.
Next to the static view, the user can interact with the visualization. When he 
hoovers over a circle of a team, only the games played by that team are shown by highliting the 
competitors (their circles) and the curved lines that connect them. This view gets 
fixated when the users clicks on a team.
He/she then automatically gets forwarded to the statistics view where the bubble 
view is still represented in the right upper corner to make sure the user keeps
an overview. The statistics view will be explained in more detail in the next section.

@@

\subsubsection{Context section}
The context section should inform the user at all times of the current selections 
that have been made. The statistics view will always be based on a specific team
and season that have been selected or that are initated. This should be clear based
on the textual information in the top left. Next to that, in the top right a small 
bubble view is shown. It indicates the team selected and how well it performed in 
the play-offs during that season, as ecplained in the bubble view section. When changing teams and/or season, this view gets 
updated as well as the textual information.

\subsubsection{Arrow section of selected statistic}
This section illustrates a change for a chosen statistic for the selected team compared with the season before. This should give a visitor more insight into why a statistic changed over time. The selected team is represented by a circle in the same way as explained in the bubble view. The change for a statistic is represented by three arrows:
\begin{itemize}
    \item Arrow on the left pointing towards the circle: indicating the influence of players who joined the team (team member current season, not previous season)
    \item Arrow on the right of the circle: indicating how the team inherent changed (team member this season and previous season)
    \item Arrow on the left pointing away from the circle: indicating the influence of players who left the team (team member previous season, not current season)
\end{itemize}
<figure of the statistics view>
The shirts next to the arrows are the shirts of players that had the biggest 
contribution to that arrow. Note that these shirts don't change in size. The box on 
the right enables the user to change between 
statistics. Next to that it informs the user of which statistic has been chosen. When 
the user hits the circle in the middle representing the teams, he is guided 
to the team view. This view will be explained in more detail in the next section.

\subsubsection{Small multiples of different statistics}
Below the arrow section small multiples are shown of multiple statistics over time. 
The line charts enables a user to identify peaks or drops that possibly influenced the 
outcome of a team is this season, following seasons or previous seasons. A user interacts with the visualization by shifting the bar indicating the current year displayed. By doing this the view gets updated and the user can see the impact on the team in the context section and why the statistic changes in the arrow section. An alternative to scroll through time is to use the timeline at the bottom of the page.

@@

\subsection{Technology}
To create the visualization, the d3 javascript framework~\cite{d3.js} was used in combination with html5~\cite{html5} and jquery~\cite{jquery}. This allows easy
access to the visualization as most modern browsers are capable of handling
these technologies.
The choice not to use the d3 framework for the entire visualization was made to
ease the layout configuration of the visualization. Instead, html was used to do
the global layout of the visualization. We opted to structure the site with multiple
divs as containers. To each container we allocate a svg with a specific visualization.
Eg. The statistic view is build out of x divs filled with different svg's created
with the d3 framework.
<sketch of how the page is structured>
When working with d3.js we encountered some obstacles:
- A synchrounous call need to be combined with synchronous calls. More specific, when a visitor uses the timeline to scroll through time, the whiping of previous visualizations was not synced with the creation of the visualisation. This resulted in whiping sections when a visualization was not created yet. Hence the view resulted in multiple figures overlapping.

\section{Lessons learned}\label{sec:discussion}
We've learned a couple of lessons during the creation of this visualization:
\begin{itemize}
    \item Find a use case as soon as possible to build your visualization around
    \item Each separate component should be a sufficient informative visualization on itself
    \item Creating a custom visualization costs time and opportunities
\end{itemize}
In what follows we will discuss our experience in more detail.

\subsection{A good use case}
At the beginning of this project we were able to design the 3 separate views 
quite quickly. Our main problem was to integrate these views as a whole. When we created drafts of two views being the bubble view and the statitics view, we suddenly noted patterns. The Golden State Warriors became NBA champion a couple of years after a great uplift in SRS score. This due to a increase inside the team and because of new players joining the team in 2009. In 
the years following, the team kept on attracting talented players and increased the inherent team score (with ups and downs) to become NBA champion in 2015. With the team view, one can even notice that the team gets stronger in center team positions during this period. The visualization however is missing at the moment if this reinforcement of the field positions
is due to the attraction of talented people or by utilizing available talent within the team.

>>>>

<Tell something more about the data in terms of tall/wide data
Tell something more about pre-processing steps (why?, did we change it over time?)
rfr to picture in 1st ppt sl 60>
\section{Related Work}\label{sec:literature}

\section{Visualization}\label{sec:visualization}
The visualization consists of three parts:
\begin{itemize}
    \item The bubble view: a compact view on the play-offs per season
    \item The statistics (zoom) view: a more detailed view on a statistic of a selected team
    \item The team view: a detailed view on how good or bad a team scores on a specific field position
\end{itemize}

<Explain that this complies with the infovis seeking mantra: Overview, zoom&filter, details on demand
Bubble = overview
statistics= details on demand
>

All three are discussed in more detail below. The three views are available to
the user on a single webpage. The user start on the bubble view and is able to
navigate to the statistics and team views by scrolling down or clicking on
teams.
%The webpage is built in such a way that it gives a hint to the user at the bottom
%or the top of a page when another view is available. By showing a different 
%background color and an arrow, we invite the visitor to explore other views. The 
%user is also at liberty to scroll through these views. However, a fixed scroll is
%implemeted so users will only end in complete views when moving around. Note that 
%the user also will be guided through all 3 views when performing actions described 
%in the section below.

\subsection{The bubble view}
This view supposes to inform the user of the results of the NBA play-offs per 
season. A circle is used to represent a team, the size its SRS score using a '?'
distribution, a stroke to illustrate it's region or medal won during regional 
competition and curved lines to inform the user which teams played against each 
other. The user has a timeline at its disposal to scroll through time. Note that
the team can be recognized by the logo inside the circle. Our initial desing filled
the circle with the red or blue depending on the region where the teams compete. 
However, the shade of colour was darker if their final ranking was higher in their
region. We've replaced this with the logo's of the teams as we thought this is more
user friendly and the extra information of the ranking doesn't outweight.
<picture of the bubble view compared with double elimination bracket>
A more popular visualization of a play-off is the team double elimination bracket.
Although this is a more common representation, we wanted to create a more compact
version by eliminating the recurring representation of a team. We also considered 
enclosed circles and a sunburst.
<figure of 2 alternatives https://thetuftersblog.wordpress.com/2016/03/07/team-ranking-play-off-infovis/>
Although the enclosed circles is a compact representation as well, we have continued 
with our own design as we believe this view is more original and innovative. We 
didn't encouter a similar view when searching for alternatives to visualize a 
play-off or championship. 
Our initial bubble view was too compact. It connected teams that played against
each other directly. When searching for alternatives to reflect on our choice we
discovered a similar visualization in a totally different context. 
<https://source.opennews.org/en-US/articles/nyts-512-paths-white-house/>
<comparison of original bubble view with white house paths>
This visualization gives a cleaner view of the competitors for a specific team
compared to our original sketch. Hence we decided to adapt our visualization. We
added an intermediate step between two teams to clearer indicate how teams competed
to become NBA champion.
Next to the static view, the user can interact with the visualization. When he 
hoovers over a circle of a team, only the games played by that team are shown by highliting the 
competitors (their circles) and the curved lines that connect them. This view gets 
fixated when the users clicks on a team.
He/she then automatically gets forwarded to the statistics view where the bubble 
view is still represented in the right upper corner to make sure the user keeps
an overview. The statistics view will be explained in more detail in the next section.
< Hierarchy = sorting = facilitate trends of winners
position have a meaning in the bubble view
when selecting circle => zoom/more detail
Bubble principle of 
- Symmetry (whole = play-off)
- conectedness arc = games played
Pre attentive characteristics
- size of circles
- hue to highligt selections or indicate increase decline arrows next view

We considered a map view, but didn't do it because geographical position has no added value
This visualization helped us discover data quality issues => cut off year 1984
>

@@

\subsubsection{Context section}
The context section should inform the user at all times of the current selections 
that have been made. The statistics view will always be based on a specific team
and season that have been selected or that are initated. This should be clear based
on the textual information in the top left. Next to that, in the top right a small 
bubble view is shown. It indicates the team selected and how well it performed in 
the play-offs during that season, as explained in the bubble view section. When changing teams and/or season, this view gets 
updated as well as the textual information.
<Explain that the context supports the short term memory of human beings rfr. Monkey Illusion>

\subsubsection{Arrow section of selected statistic}
This section illustrates a change for a chosen statistic for the selected team compared with the season before. This should give a visitor more insight into why a statistic changed over time. The selected team is represented by a circle in the same way as explained in the bubble view. The change for a statistic is represented by three arrows:
\begin{itemize}
    \item Arrow on the left pointing towards the circle: indicating the influence of players who joined the team (team member current season, not previous season)
    \item Arrow on the right of the circle: indicating how the team inherent changed (team member this season and previous season)
    \item Arrow on the left pointing away from the circle: indicating the influence of players who left the team (team member previous season, not current season)
\end{itemize}
<figure of the statistics view>
The shirts next to the arrows are the shirts of players that had the biggest 
contribution to that arrow. Note that these shirts don't change in size. The box on 
the right enables the user to change between 
statistics. Next to that it informs the user of which statistic has been chosen. When 
the user hits the circle in the middle representing the teams, he is guided 
to the team view. This view will be explained in more detail in the next section.
<filtering: 
- time & team
- based on exploration (small multiples)
- dynamically: changes impact multiple parts>

\subsubsection{Small multiples of different statistics}
Below the arrow section small multiples are shown of multiple statistics over time. 
The line charts enables a user to identify peaks or drops that possibly influenced the 
outcome of a team is this season, following seasons or previous seasons. A user interacts with the visualization by shifting the bar indicating the current year displayed. By doing this the view gets updated and the user can see the impact on the team in the context section and why the statistic changes in the arrow section. An alternative to scroll through time is to use the timeline at the bottom of the page.
<this vis helps us comparre statistics over time and relations between them
eg. srs score & play-off ranking>

@@

\section{Possible improvements}
As with most projects, the result is never 'finished', meaning that there are always improvements
possible but due to timing or resource constraints the haven't all been implemented. In what 
follows, we suggest a number of features that might improve the visualization:
\begin{itemize}
    \item Smooth transitions between states could improve the user experience. For example when changing
    year, a transformation of the arrow in the statistics view would help the user see a difference between years.
    At the moment the arrows change at the blink of an eye, making it difficult for us humans with a short term memory
    to make the comparison.
    \item To further support our short term memory we could also 'save' the previous state. Eg. in the field view we could
    visualize ghost shirts to indicate which players left the team or position and highlight players that joined the team. 
    This would help the user see how the team changes over time.
    \item At the moment no attention has been payed to the process & provenance of the exploration. A user thus needs to 
    memorize his action to be able to share or reproduce his actions. This could be improved by adding URL parameters. This
    way a user at least can share or save a state of an analysis. One step further is to provide a play-feature. The visitor could 
    them share his exploration and replay it. A simple implemention would ask the visitor for a start date and an end date and
    would then 'play' the changes of a particular team and characteristic over time.
\end{itemize}

\subsection{Technology}
To create the visualization, the d3 javascript framework~\cite{d3.js} was used in combination with html5~\cite{html5} and jquery~\cite{jquery}.  Note that D3 is based on HTML, CSS, Javascript and SVG.
Next to that,the use of D3 was a requirement for the project. This allows easy
access to the visualization as most modern browsers are capable of handling
these webstandards. 
The choice not to use the d3 framework for the entire visualization was made to
ease the layout configuration of the visualization. Instead, html was used to do
the global layout of the visualization. We opted to structure the site with multiple
divs as containers. To each container we allocate a svg with a specific visualization.
Eg. The statistic view is build out of x divs filled with different svg's created
with the d3 framework.
<sketch of how the page is structured>
When working with d3.js we encountered some obstacles:
- A synchrounous call need to be combined with synchronous calls. More specific, when a visitor uses the timeline to scroll through time, the whiping of previous visualizations was not synced with the creation of the visualisation. This resulted in whiping sections when a visualization was not created yet. Hence the view resulted in multiple figures overlapping.

\section{Lessons learned}\label{sec:discussion}
We've learned a couple of lessons during the creation of this visualization:
\begin{itemize}
    \item Find a story to tell with your visualization
    \item Each separate component should be a sufficient informative visualization on itself
    \item Creating a custom visualization costs time and opportunities
\end{itemize}
In what follows we will discuss our experience in more detail.

\subsection{Story telling}
At the beginning of this project we were able to design the 3 separate views 
quite quickly. Our main problem was to integrate these views as a whole. When we created drafts of two views being the bubble view and the statitics view, we suddenly noted patterns. The Golden State Warriors became NBA champion a couple of years after a great uplift in SRS score. This due to a increase inside the team and because of new players joining the team in 2009. In 
the years following, the team kept on attracting talented players and increased the inherent team score (with ups and downs) to become NBA champion in 2015. With the team view, one can even notice that the team gets stronger in center team positions during this period. The visualization however is missing at the moment if this reinforcement of the field positions
is due to the attraction of talented people or by utilizing available talent within the team. Rfr to possible improvements. Being able to visualize this story, enabled us to support visistors in searching their own stories. They are able to explore the data for patters. On top of that, we provide the insight on how the changes were impacted by changes in teamplayers.
